---
title: "BiocAlphaMissense: Bioconductor interfaces to AlphaMissense pathogenicity findings"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{BiocAlphaMissense: Bioconductor interfaces to AlphaMissense pathogenicity findings}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

AlphaMissense ([Science 2023](https://www.science.org/doi/epdf/10.1126/science.adg7492))
is a machine learning procedure for inferring pathogenicity of single nucleotide
substitutions in the human proteome.  Text files are distributed by
the authors, and these have been tabix-indexed for use in this
package.  Note that the pathogenicity data are produced under a
license that forbids commercial use.

The package described here can currently be installed using
```
BiocManager::install("vjcitn/BiocAlphaMissense")
```

# A quick look at single-nucleotide data

The following code draws the first 10 records and
places them in a data.frame.  Note that 600 MB of
bgzipped text and a tabix index will be retrieved
and cached on first attempt.

```{r lk1, message=FALSE}
library(BiocAlphaMissense)
library(Rsamtools)
amis_txf = get_alphamis_txf(build="hg38")
amis_txf
yieldSize(amis_txf) = 10L
df10 = read.delim(text=scanTabix(amis_txf)[[1]], h=FALSE)
data(amnames)
names(df10) = amnames
df10
```

# Checking against GWAS hits: asthma example

We used the EBI GWAS catalog and searched for "sthma"
in the `MAPPED_TRAIT` field.  The resulting
records are in a GRanges instance, accessible via `data(amentgr)`.

```{r lkda}
data(amentgr)
length(amentgr)
amentgr[1:3,c("STRONGEST.SNP.RISK.ALLELE", "PUBMEDID")]
```

The coincidence of these GWAS hits with AlphaMissense
results (`amis_txf` created above) can be computed quickly,.

```{r doint, message=FALSE}
library(Rsamtools)
library(GenomeInfoDb)
seqlevels(amentgr) = paste0("chr", seqlevels(amentgr)) # if off line
yieldSize(amis_txf) = NA_integer_
lk = scanTabix(amis_txf, param=amentgr)
```

Some of the positions with GWAS hits don't correspond to
results from AlphaMissense.  These positions are
empty components of the list returned by scanTabx.
They are removed and the scanTabix results are converted
to a data.frame.
```{r doconv}
ok = which(vapply(lk, function(x)length(x)>0, logical(1)))
cc = c("character", "character", "character", "character", "character", 
"character", "character", "character", "numeric", "character")
intsect = lapply(lk[ok], 
   function(x) read.delim(text=x, h=FALSE, colClasses=cc))
intsect_df = do.call(rbind, intsect)
data(amnames)
names(intsect_df) = amnames
head(intsect_df,3)
```

At this point, `intsect_df` is the collection of _all_
substitution scores at asthma GWAS hits.
We focus on the substitutions reported as risk alleles.

```{r lkrisk}
amentgr$RISK_ALLELE = gsub("(.*-)", "", amentgr$STRONGEST.SNP.RISK.ALLELE)
```

To join the missense classes with the coincident GWAS hits
we build a "key" for each table and then join.

```{r dokey}
intsect_df$key = with(intsect_df, paste(CHROM, POS, ALT, sep=":"))
ament_df = as.data.frame(amentgr)
ament_df$key = with(ament_df, paste(seqnames, CHR_POS, RISK_ALLELE, sep=":"))
ia = dplyr::inner_join(intsect_df, ament_df, by="key", relationship = "many-to-many")
iau = ia[!duplicated(ia$key),]
table(iau$am_class)
```

Some of the information about the likely pathogenic substitutions
that have been identified as hits for asthma
is collected here:

```{r lktab}
iau |> 
  dplyr::filter(am_class == "likely_pathogenic") |> 
  dplyr::select(CHR_ID, CHR_POS, STRONGEST.SNP.RISK.ALLELE, 
  OR.or.BETA, MAPPED_TRAIT, MAPPED_GENE, am_class)
```

